---
title    : ECON 6931 - R Programming (Code Supplement)
output:
  html_document:
    toc: true
    theme: united
---

<h1>Week 1 Programming Supplement</h1>

This document contains code that were too long to display in the week's slides. Unlike code in the slides, all examples below are executed so you can see the values.

You should use these programming supplements to follow along and try executing the example code yourself.

<h2>Variables and Namespaces</h2>

When you execute a statement like `x <- 5` in R, you are creating an **object** in memory which holds the numeric **value** 5 and is referenced by the **variable name** "x".

If you later ask R to do something like `y <- x + 2`, it will search sequentially through a series of **namespaces** until it finds a variable called "x". Namespaces can be thoguht of as collections of labels pointing to places in memory. You can use R's `search` command to examine the ordered list of namespaces that R will search in for variables.

```{r namespaces, eval = TRUE}
# Check the search path of namespaces
search()

# use ls() to list the objects in one of those namespaces
ls("package:stats")
```

<h2>Dollarstore Calculator Math</h2>

```{r simpleMath1, eval = TRUE, echo = TRUE}
# Addition with "+"
4 + 5

# Subtraction with "-"
100 - 99

# Multiplication with "*"
4 * 5

# Division with "/"
15 / 3

# Exponentiation with "^"
2^3

# Order of Operations
4 * 5 + 5 / 5

# Control with parentheses
4 * (5 + 5) / 5
```

<h2>Controlling Program Flow: If-Else</h2>

Soon after you start writing code (in any language), you'll inevitably find yourself saying "I only want to do this thing if certain conditions are met". This type of logic is expressed using [if-else syntax](https://en.wikipedia.org/wiki/Conditional_(computer_programming))

```{r ifElse1, eval = TRUE, echo = TRUE, warning = FALSE, message = FALSE}
library(lubridate)
library(gaussfacts)

# Set some variable
DAY <- as.character(wday(now(), label = TRUE, abbr = FALSE))
print(DAY)

# If it's Monday, print 2 Gauss facts. Otherwise, print 1
if (DAY == "MONDAY"){
    gaussfact()
    gaussfact()
} else {
    gaussfact()
}
```

What if you want to express more than two possible outcomes? For this, we could use R's `else if` construct to nest conditions. Note that conditional blocks can have any number of "else if" statements, but only one "else" block.

```{r ifElse2, eval = TRUE, echo = TRUE}
# Try to think through what this will do before you run it yourself:
if (4 > 5){
    print("3")
} else if (6 <= (5/10)) {
    print("1")
} else if (4 + 4 + 4 == 12.0001) {
    print("4")
} else {
    print("2")
}
```

<h2>Controlling Program Flow: For Loops</h2>

One of the most powerful characteristics of general purpose programming languages is their ability to automate repetitive tasks. When you know that you want to do something a fixed number of times (say, squaring each item in a vector), you can use a `for` loop.

```{r forLoop, eval = TRUE, echo = TRUE}
# Create a vector
x <- c(1,4,6)

# Print the square of each element one at a time
print(1^2)
print(4^2)
print(6^2)

# BETTER: Loop over the vector and print the square of each element
for (some_number in x){
    print(some_number^2)
}
```

<h2>Controlling Program Flow: While Loops</h2>

For loops are suitable for many applications, but they can be too restrictive in some cases. For example, imagine that you are writing a simple movie search engine and you want to tell R "look through an alphabetized list of movies and tell me if you find Apocalypse Now". A `for` loop can certainly do this, but it will keep running over ALL movies...long after it finds Ace Ventura! This is a great place to use a `while` loop.

Here is the `for` loop implementation:

```{r whileLoop1, eval = TRUE, echo = TRUE}
movie_list <- c("ace ventura", "apocalypse now", "return of the jedi", "v for vendetta", "zoolander")
MOVIE_TO_SEARCH_FOR <- "apocalypse now"

# Naive for loop implementation
i <- 1
for (movie in movie_list) {
    if (movie == MOVIE_TO_SEARCH_FOR) {
        print(paste0(i, ": found it!"))
    } else {
        print(paste0(i, ": not found"))
    }
    i <- i + 1
}
```

And here is the `while` loop implementation. Notice that this one stops checking once it finds what it wants.

```{r whileLoop2, eval = TRUE, echo = TRUE}
movie_list <- c("ace ventura", "apocalypse now", "return of the jedi", "v for vendetta", "zoolander")
MOVIE_TO_SEARCH_FOR <- "apocalypse now"

# Faster while loop implementation
KEEP_SEARCHING = TRUE
i = 1
while (KEEP_SEARCHING){
    
    # Check this element
    if (movie_list[i] == MOVIE_TO_SEARCH_FOR) {
        print(paste0(i, ": found it!"))
        KEEP_SEARCHING = FALSE
    } else {
        print(paste0(i, ": not found"))
    }
    
    # If we've reached the end, break out. Otherwise, increment the counter
    if (i == length(movie_list)){
        print("Done searching. This movie isn't in the list")
    } else {
        i <- i + 1
    }
}
```

<h2>Required Arguments</h2>

- R functions take 0 or more arguments...basically named variables that the function uses to do it's work
- Take a look at `?sqrt`. You'll see that it takes one argument, named `x`. You can pass any vector of numeric values to this argument and `sqrt` will return the square root of each element
- In this case, we'd say `x` is a *required argument* of `sqrt`

```{r requiredArgs, eval = TRUE, echo = TRUE, error = TRUE}
# Take the square root of a vector of numbers
sqrt(x = c(1,4,9,16,25))

# Note that calling this function without the argument will throw an error!
sqrt()
```

<h2>Default Argument Values</h2>

- For more complicated functions, passing values to each argument can get burdensome
- To handle this, R allows function authors to specify *default arguments*. These are values that certain arguments will take automatically unless you decide to overwrite them
- Example: look at `?rnorm`. You'll see that this function's signature reads `rnorm(n, mean = 0, sd = 1)`. 

```{r normalDraw}
# 100 random draws from a normal distribution w/ mean 0 and standard deviation 1
rand_nums <- rnorm(n = 100)

# 100 random draws from a normal distribution w/ mean 4.5 and standard deviation 1
rand_nums <- rnorm(n = 100, mean = 4.5)
```

<h2>Vectors</h2>

- Because R was designed for use with statistics, most of its operations are [vectorized](http://www.cs.cornell.edu/courses/cs1112/2016sp/Exams/exam2/vectorizedCode.pdf)
- You can create vectors a few ways:

```{r vectors_pt1, eval = TRUE}
# Ordered sequence of integers
1:5

# Counting by 2s
seq(from = 0, to = 14, by = 2)

# Replicate the same values
rep(TRUE, 6)

# Concatenate multiple values with the "c" operator
c("all", "of", "the", "lights")

# Watch out! Mixing types wil lead to silent coercion
c(1, TRUE, "hellos")

# Some functions, when applied over a vector, return a single value
is.numeric(rnorm(100))

# Others will return a vector of results
is.na(c(1, 5, 10, NA, 8))

# Vectors can be named
batting_avg <- c(youkilis = 0.300, ortiz = 0.355, nixon = 0.285)

# You can combine two vectors with c()
x <- c("a", "b", "c")
y <- c("1", "2", "3")
c(x, y)
```

<h2>Lists</h2>

Vectors are the first multi-item data structure all R programmers learn. Soon, though, you may find yourself frustrated with the fact that they can only hold a single type. To handle casses where you want to package multiple types (and even multiple objects!) together, we will turn to a data structure called a `list`.

|Capabilities                        |Vectors |Lists   |
|:----------------------------------:|:------:|:------:|
|Optional use of named elements      |&#x2714;|&#x2714;|
|Support math operations like mean() |&#x2714;|        |
|Hold multiple types                 |        |&#x2714;|
|Hold multiple objects               |        |&#x2714;|


```{r lists_pt1, eval = TRUE}
# Create a list with list()
myList <- list(a = 1, b = rep(TRUE, 10), x = c("shakezoola", "mic", "rulah"))

# Examine it with str()
str(myList)
```

We saw earlier how you can use `for` loops to walk over a vector and do something with each element. You can certainly do the same thing with a list, but it can be a bit tedious.

```{r listLoop}
# Create a list
studentList <- list(kanye = c(80, 90, 100), talib = c(95, 85, 99), common = c(100, 100, 99))

# Get average grades with a for loop
grades <- vector(mode = "list", length = 3)
for (i in 1:length(studentList)){
    grades[[i]] <- mean(studentList[[i]])
}
```

Luckily, R provides a built-in function called `lapply` to handle this more expressively. We'll cover R's [apply family of functions](https://www.datacamp.com/community/tutorials/r-tutorial-apply-family#gs.ER5hbLU) later in the course. Here's a sneak peek:

```{r listLapply}
# Better way with lapply
grades <- lapply(studentList, mean)
```

<h2>Data Frames</h2>

Vectors and lists are crucial data structures in R, but you may find that they're difficult to work with in model training and other data science tasks. It is now time to introduce a third foundational data structure: the data frame.

Data frames are tables of data. Each column of a dataframe can be a different type, but all values within a column must be the same type. 

```{r dataFrame}
# Create a dataframe!
myDF <- data.frame(time_period  = c(1, 2, 3, 4, 5),
                   temperature  = c(25.6, 38.7, 31.4, 40.0, 29.20),
                   station      = c("A", "B", "A", "A", "B"),
                   is_gov_owned = c(TRUE, FALSE, TRUE, TRUE, FALSE)
                   )

# Check out the structure of this thing
str(myDF)
```

R comes with some sample data sets you can experiment with. Let's load the `mtcars` data.frame and test out some new commands!

```{r dateFrame2}
# Load the mtcars dataframe
data("mtcars")

# Check out its structure
str(mtcars)

# View the top 10 rows
head(mtcars, n = 10) # could use "tail" for the bottom 10

# Find all the unique values of "cyl" (the number of engine cylinders)
unique(mtcars$cyl)
```

We will be working with data frames throughout this class, given their importance to data science and statistics. A more thorough treatment of this data structure will be given in Week 2. For now, I'd like to introduce you to one more looping function: `apply`.

`apply` allows you to loop over the rows or columns of a data frame and execute an arbitrary function. The code below holds some examples of what can be accomplished with `apply`.

```{r apply_on_a_DF, eval = TRUE}
# Get the mean of each column
apply(mtcars, MARGIN = 2, FUN = mean)

# Get the mean of each row (nonsensical, just illustrating
apply(mtcars, MARGIN = 1, FUN = mean)
```